The World Bank, Women, Business and the Law

### 2024-02-21 outputs/paper/paper.qmd
updated references

Despite the many decades' worth of progress in gender equality throughout the world, one of the enduring types of gender discrimination worldwide is in the eyes of the law. Though much progress has been made in western, high-income countries, there remains vast swaths of the world where women may not vote, work, or participate in the same activities as men do, due to the presence of gendered laws. To trace the progression of economic opportunities of women in the world's 190 countries, the World Bank constructed their "Women, Business, and the Law" (WBL) dataset, which Hyland et. al make of primary interest in their paper "Gendered Laws and Women in the Workforce". In Sections 2 and 3, we analyze and replicate some of the key graphs and insights from their paper. Building on their work in Sections 4 and 5, we  call into question the data collection methods of the World Bank, finding data from Statistics Canada that conflict with the data [or whatever Sehar finds], outlining the importance that accurate data have in being able to fully understand the complex picture of women and the law throughout the world.

# Data {#sec-data}

@@

## Source {#subsec1-data}

@@

	Mobility: Examines constraints on women’s freedom of movement. 
	Workplace: Analyzes laws affecting a woman’s decision to work
	Pay: Measures laws and regulations affecting a woman’s pay
	Marriage: Assesses constraints related to marriage
	Parenthood: Examines laws that affect women’s work after having children
	Entrepreneurship: Assesses constraints to women starting and running a business
	Assets: covers property ownership rights, inheritance rights (both for children and surviving spouses), authority of assets during marriage, and valuation of non-monetary contributions. 
	Pension: captures the equalization of retirement ages (with full and partial benefits as well as the mandatory retirement age) and whether periods of absence from employment due to childcare are accounted for in pension benefits.

	Indicator-level scores are obtained by calculating the unweighted average of the questions within that indicator and scaling the result to 100. Overall scores are then calculated by taking the average of each indicator, with 100 representing the highest possible score. Each binary question either contributes 20 or 25 points, depending on the indicator. The indicators and index mainly capture the effect labour laws have on the employment outcomes of women in the labour force. It does not take into account laws affecting their acquisition of human capital (e.g. required education laws) nor laws affecting reproductive rights, for instance, which Hyland et al. point to as an issue but concede would be difficult to quantify easily. The World Bank's methodology was designed as an easily replicable measure of the legal environment for women as entrepreneurs and employees, which, for a dataset of it's size and breadth, is comprehensive. In their paper, they then use the dataset to present 'stylized facts' which either unearth new facts about gendered laws or are quantifications of what they write were previously known qualitatively.

	Figures 1A, 1B, and 2 were replicated from the original paper.

@@

R [cite] and it's tidyverse [cite] and dplyr [cite] packages were primarily used when analysing the WBL dataset. Like Hyland, this dataset will be of main focus in our analysis. Other datasets that helped their paper, such as Wage Gap data from the International Labour Organization and the OECD (Organization for Economic Cooperation and Development), and World Development Indicators data is not important to our analysis.

@@

# Analysing WBL Indicators from a Canadian viewpoint

## Pay

The first indicator to be scrutinized is Pay, particularly the question that focuses on equal remuneration regardless of gender. Notably, the legal backing for an indicator representative of Canada is one that references only one province, Ontario. Nevertheless, this section will utilize the LFS Survey from Statistics Canada to evaluate if Canada’s perfect rating doesn’t hold up solely on paper.

```{r, results='asis', echo=FALSE}
library(kableExtra)

# Create a data frame for the table
pay <- data.frame(
  Indicator = "Pay",
  Score = 100,
  Question = "“Does the law mandate equal remuneration for work of equal value?”",
  Legal_Basis = "Ontario Pay Equity Act, Sec. 6",
  Achievement_of_pay_equity = "6 (1) For the purposes of this Act, pay equity is achieved under the job-to-job method of comparison when the job rate for the female job class that is the subject of the comparison is at least equal to the job rate for a male job class in the same establishment where the work performed in the two job classes is of equal or comparable value. R.S.O. 1990, c. P.7, s. 6 (1); 1993, c. 4, s. 4 (1).",
  Relevant_article = "The gender wage gap in Canada: 1998 to 2018 C Labour Statistics: Research Papers by Rachelle Pelletier, Martha Patterson and Melissa Moyser",
  Stylized_fact = "Women still earn 87 cents for every dollar a man earns in Canada",
  Link_to_Article = "[Read more](https://www150.statcan.gc.ca/n1/en/pub/75-004-m/75-004-m2019004-eng.pdf?st=szPFyv6s)",
  Database = "Annualized data from the Labour Force Survey (LFS), a monthly household survey conducted by Statistics Canada.",
  Additional_info = "56,000 households, no aboriginal settlements or full-time Canadian Armed Force members, remote households in low population density areas"
)

@@

# Create a table using kableExtra
  kable(pay, format = "latex", caption = "Summary of findings -Pay Indicator") %>%
  column_spec(2, width = "27em")

The LFS survey reveals that in 2018, female employees aged 25 to 34 earned $4.13 less per hour compared to their male counterparts, equaling 87 cents on the dollar. In the decade between 1988 and 2018, the gap has reduced by merely $1.04. 
Given that women in Canada are now better educated on average than men, acquiring bachelor’s degrees at a faster pace, and have diversified their fields of study, particularly in STEM, the persistence of this gap warrants a closer look.

@@

```{r, results='asis', echo=FALSE, fig.cap="Explanation of the Gender Wage Gap, 1998 to 2018"}

library(kableExtra)

# Create a data frame for the table
wagegap <- data.frame(
  Category = c("Total explained portion (sum of effects of variables below)", "Human capital", "Education", 
               "Job tenure", "Job attributes", "Part-time", "Public sector", "Union", "Firm size", 
               "Occupation and industry", "Occupation", "Industry", "Demographics", "Age", 
               "Province", "Children", "Marital status", "Total unexplained portion"),
  'In 1998' = c(28.0, 1.8, -0.5, 2.3, 10.1, 8.9, -0.6, 0.7, 1.1, 18.3, 1.8, 16.5, -2.2, -0.4, -0.1, -0.3, -1.4, 72.0),
  'In 2018' = c(36.6, -6.1, -4.8, -1.3, -0.7, 9.2, -5.3, -3.4, -1.1, 44.8, 5.1, 39.7, -1.6, -0.3, 1.0, -0.8, -1.5, 63.4)
)

@@

# Create a table using kableExtra
kable(wagegap, format = "latex", booktabs = TRUE, escape = FALSE) %>%
 kable_styling() %>%
#  add_header_above(c("Category" = 1, "1998" = 1, "2018" = 1)) %>%
  row_spec(18, bold = TRUE)

@@

## Workplace {#sec-point2}

@@

```{r, results='asis', echo=FALSE}
library(kableExtra)

# Create a data frame for the table
workplace <- data.frame(
  Indicator = "Workplace",
  Score = 100,
  Question = "“Is there legislation on sexual harassment in employment??”",
  Legal_Basis = "Canadian Human Rights Act, Sec. 14(1) and (2); Ontario Occupational Health and Safety Act, Sec. 1 and 32.0.1; Ontario Human Rights Code, Sec. 7(2) and 7(3)",
  Code = "14 (1) It is a discriminatory practice, in matters related to employment,
to harass an individual on a prohibited ground of discrimination.Without limiting the generality of subsection 1, sexual harassment shall, for the purposes of that subsection, be deemed to be harassment on a prohibited ground of discrimination.",
  Relevant_article = "Gender Results Framework: A new data table on workplace harassment",
  Stylized_fact = "Women still earn 87 cents for every dollar a man earns in Canada",
  Link_to_Article = "[Read more]https://www150.statcan.gc.ca/n1/daily-quotidien/240212/dq240212a-eng.htm",
  Database = "Data from 2020 Survey on Sexual Misconduct at Work (SSMW) and Gender Results Framework",
  Additional_info = "56,000 households, no aboriginal settlements or full-time Canadian Armed Force members, remote households in low population density areas"
)

@@

# Replace underscores with spaces in column names
colnames(workplace) <- gsub("_", " ", colnames(workplace))

# Create a table using kableExtra
  kable(workplace, format = "latex", caption = "Summary of findings -Pay Indicator") %>%
  column_spec(2, width = "27em")

@@

The reproduced table from the 2020 Survey on Sexual Misconduct at Work outlines the proportions of employees reporting inappropriate sexualized behavior, discriminatory behavior, and sexual assault in their work environments. Notably, the data highlights the heightened vulnerability of individuals in the 25 to 34 age group, where 57% of females report experiencing sexualized behaviors. Within this age group, women are twice as likely as men to face discrimination and three times as likely to experience sexual assault.
```{r}
# Install and load necessary packages if not already installed
 # install.packages("tidyverse")

# Install and load necessary packages if not already installed
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

@@

# Your provided data
data <- read.table(text = '"Group" "M - inappropriate sexualized behaviours" "M - Discrimination" "M - SA" "F - inappropriate sexualized behaviours" "F - Discrimination" "F - SA"
"Total, all people" 29 9 3 44 20 13
"Age 15 to 24" 27 12 0 48 25 7
"Age 25 to 34" 37 13 5 57 27 16
"Age 35 to 44" 34 11 3 41 20 12
"Age 45 to 54" 28 8 3 42 18 12
"Age 55 and older" 24 6 2 39 16 15', header = TRUE)

@@

# Create a sideways stacked bar plot
ggplot(data_long, aes(x = Percentage, y = Group, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("#FF9999", "#FF6666", "#FF3333", "#66B2FF", "#3366FF", "#000099"), name = "Category") +
  labs(title = "Sideways Stacked Bar Plot",
       x = "Percentage",
       y = "Group") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.margin = margin(t = 0, unit = "pt")
)

@@

## Weaknesses and next steps

@@

::: {#refs}
:::

@@

# Appendix {-}

>>>>

Despite the many decades' worth of progress in gender equality throughout the world, one of the enduring types of gender discrimination worldwide is in the eyes of the law. Though much progress has been made in western, high-income countries, there remains vast swaths of the world where women may not vote, work, or participate in the same activities as men do, due to the presence of gendered laws. To trace the progression of economic opportunities of women in the world's 190 countries, the World Bank constructed their "Women, Business, and the Law" (WBL) dataset, which Hyland et. al make of primary interest in their paper "Gendered Laws and Women in the Workforce". In Sections 2 and 3, we analyze and replicate some of the key graphs and insights from their paper. Building on their work in Sections 4 and 5, we call into question the data collection methods of the World Bank, finding data from Statistics Canada that conflict with the data \[or whatever Sehar finds\], outlining the importance that accurate data have in being able to fully understand the complex picture of women and the law throughout the world.

# Data

## Source {#sec-data}

@@

`R` [@citeR] and it's `tidyverse` [@tidyverse] and `dplyr` [@dplyr] packages were primarily used when analysing the WBL dataset. Like Hyland, this WBL dataset will be of primary focus in our reproduction. Other datasets that informed the original paper and were cross-references, such as Wage Gap data from the International Labour Organization and the OECD (Organization for Economic Cooperation and Development), and World Development Indicators data are not relevant to the areas that will be developed upon, and so irrelevant to the subsequent analysis. For the secondary part of this analysis, Surveys conducted by Statistics Canada will be utilised.

## Features {#subsec1-data}

@@

• Mobility: Examines constraints on women’s freedom of movement.\
• Workplace: Analyzes laws affecting a woman’s decision to work.\
• Pay: Measures laws and regulations affecting a woman’s pay.\
• Marriage: Assesses constraints related to marriage\
• Parenthood: Examines laws that affect women’s work after having children.\
• Entrepreneurship: Assesses constraints to women starting and running a business.\
• Assets: covers property ownership rights, inheritance rights (both for children and surviving spouses), authority of assets during marriage, and valuation of non-monetary contributions.\
• Pension: captures the equalization of retirement ages (with full and partial benefits as well as the mandatory retirement age) and whether periods of absence from employment due to childcare are accounted for in pension benefits.

Indicator-level scores are obtained by calculating the unweighted average of the questions within that indicator and scaling the result to 100. Overall scores are then calculated by taking the average of each indicator, with 100 representing the highest possible score. Each binary question either contributes 20 or 25 points, depending on the indicator. The indicators and index mainly capture the effect labour laws have on the employment outcomes of women in the labour force. It does not take into account laws affecting their acquisition of human capital (e.g. required education laws) nor laws affecting reproductive rights, for instance, which Hyland et al. point to as an issue but concede would be difficult to quantify easily. The World Bank's methodology was designed as an easily replicable measure of the legal environment for women as entrepreneurs and employees, which, for a dataset of it's size and breadth, is comprehensive. In their paper, they then use the dataset to present 'stylized facts' which either unearth new facts about gendered laws or are quantifications of what they write were previously known qualitatively. Figures 1A, 1B, and 2 int the following sections were replicated from the original paper.

## Key Results

@@

@@

### Pay

The first indicator to be scrutinized is Pay, particularly the question that focuses on equal remuneration regardless of gender. Notably, the legal backing for an indicator representative of Canada is one that references only one province, Ontario. Nevertheless, this section will utilize the LFS Survey from Statistics Canada to evaluate if Canada’s perfect rating doesn’t hold up solely on paper.  
```{r canada-tbl-1}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
library(kableExtra)

# Create a data frame for the table
pay <- data.frame(
  Indicator = "Pay",
  Score = 100,
  Question = "“Does the law mandate equal remuneration for work of equal value?”",
  Legal_Basis = "Ontario Pay Equity Act, Sec. 6",
  Citation = "For the purposes of this Act, pay equity is achieved under the job-to-job method of comparison when the job rate for the female job class that is the subject of the comparison is at least equal to the job rate for a male job class in the same establishment where the work performed in the two job classes is of equal or comparable value.",
  Article = "The gender wage gap in Canada: 1998 to 2018 Labour Statistics: Research Papers by Rachelle Pelletier, Martha Patterson and Melissa Moyser",
  Stylized_fact = "Women still earn 87 cents for every dollar a man earns in Canada",
  Link = "[Read more](https://www150.statcan.gc.ca/n1/en/pub/75-004-m/75-004-m2019004-eng.pdf?st=szPFyv6s)",
  Database = "Annualized data from the Labour Force Survey (LFS), a monthly household survey conducted by Statistics Canada.",
  Additional = "56,000 households, no aboriginal settlements or full-time Canadian Armed Force members, remote households in low population density areas"
)

@@

# Create a table using kableExtra
  kable(pay, format = "latex", caption = "Summary of findings -Pay Indicator") %>%
  column_spec(2, width = "29em")
```

The LFS survey reveals that in 2018, female employees aged 25 to 34 earned 4.13 dollars less per hour compared to their male counterparts, equaling 87 cents on the dollar. In the decade between 1988 and 2018, the gap has reduced by merely \$1.04. Given that women in Canada are now better educated on average than men, acquiring bachelor’s degrees at a faster pace, and have diversified their fields of study, particularly in STEM, the persistence of this gap warrants a closer look.

@@

```{r canada-tbl-2}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
library(kableExtra)

# Create a data frame for the table
wagegap <- data.frame(
  Category = c("Total explained portion (sum of effects of variables below)", "Human capital", "Education", 
               "Job tenure", "Job attributes", "Part-time", "Public sector", "Union", "Firm size", 
               "Occupation and industry", "Occupation", "Industry", "Demographics", "Age", 
               "Province", "Children", "Marital status", "Total unexplained portion"),
  '1998' = c(28.0, 1.8, -0.5, 2.3, 10.1, 8.9, -0.6, 0.7, 1.1, 18.3, 1.8, 16.5, -2.2, -0.4, -0.1, -0.3, -1.4, 72.0),
  '2018' = c(36.6, -6.1, -4.8, -1.3, -0.7, 9.2, -5.3, -3.4, -1.1, 44.8, 5.1, 39.7, -1.6, -0.3, 1.0, -0.8, -1.5, 63.4)
)

colnames(wagegap) <- c("Category", "1998", "2018")

@@

# Create a table using kableExtra
kable(wagegap, format = "latex", booktabs = TRUE, escape = FALSE, caption = "Explanation of the Gender Wage Gap, 1998 to 2018") %>%
 kable_styling() %>%
#  add_header_above(c("Category" = 1, "1998" = 1, "2018" = 1)) %>%

  row_spec(18, bold = TRUE)

@@

### Workplace {#sec-point2}

@@

```{r canada-tbl-3}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
library(kableExtra)

# Create a data frame for the table
workplace <- data.frame(
  Indicator = "Workplace",
  Score = 100,
  Question = "“Is there legislation on sexual harassment in employment?”",
  Legal_Basis = "Canadian Human Rights Act, Sec. 14(1) and (2)",
  Citation = "It is a discriminatory practice,in matters related to employment,to harass an individual on a prohibited ground of discrimination.",
  Article = "Gender Results Framework: A new data table on workplace harassment",
  Stylized_fact = "Women still earn 87 cents for every dollar a man earns in Canada",
  Link = "[Read more](https://www150.statcan.gc.ca/n1/daily-quotidien/240212/dq240212a-eng.htm)",
  Database = "Data from 2020 Survey on Sexual Misconduct at Work (SSMW) and Gender Results Framework."
)

@@

# Replace underscores with spaces in column names
colnames(workplace) <- gsub("_", " ", colnames(workplace))
# Create a table using kableExtra
  kable(workplace, format = "latex", caption = "Summary of findings -Workplace Indicator") %>%

  column_spec(2, width = "29em")
```

@@

The reproduced table from the 2020 Survey on Sexual Misconduct at Work outlines the proportions of employees reporting inappropriate sexualized behavior, discriminatory behavior, and sexual assault in their work environments. Notably, the data highlights the heightened vulnerability of individuals in the 25 to 34 age group, where 57% of females report experiencing sexualized behaviors. Within this age group, women are twice as likely as men to face discrimination and three times as likely to experience sexual assault.

```{r canada-fig-1}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center

# Install and load necessary packages if not already installed
 # install.packages("tidyverse")

@@

#  provided data
data <- read.table(text = '"Group" "M - inappropriate sexualized behaviours" "M - Discrimination" "M - SA" "F - inappropriate sexualized behaviours" "F - Discrimination" "F - SA"
"Total, all people" 29 9 3 44 20 13
"Age 15 to 24" 27 12 0 48 25 7
"Age 25 to 34" 37 13 5 57 27 16
"Age 35 to 44" 34 11 3 41 20 12
"Age 45 to 54" 28 8 3 42 18 12
"Age 55 and older" 24 6 2 39 16 15', header = TRUE)

@@

# Create a sideways stacked bar plot
ggplot(data_long, aes(x = Percentage, y = Group, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge", width = 1) +
  scale_fill_manual(values = c("#FF9999", "#FF6666", "#FF3333", "#66B2FF", "#3366FF", "#000099"), name = "Category") +
  labs(title = "Inappropriate workplace behaviour ",
       x = "Percentage affected ",
       y = "Age Group") +

  theme_minimal() +
  theme(
 #   axis.text.y = element_text(angle = 0, hjust = 2),
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.box = "horizontal",
)

@@

### Parenthood

```{r canada-tbl-4}
#| message: false
#| echo: false
#| warning: false
#| fig-align: center
library(kableExtra)

# Create a data frame for the table
maternal <- data.frame(
  Indicator = "Parenthood",
  Score = 100,
  Question = "“Is there paid parental leave?”",
  Legal_Basis = "Employment Insurance Act, Art. 12(4); Employment Insurance Act, Secs. 12(4.01) and 23",
  Citation = "The maximum number of weeks for which benefits may be paid for a single pregnancy is 15",
  Article = "2021 MATERNITY LEAVE EXPERIENCE REPORT, Moms At Work",
  Stylized_fact = "One in every 3 mothers returning to the workplace experienced discrimination",
  Link = "[Read more](https://thisismomsatwork.com/wp-content/uploads/edd/2021/05/2021-Maternity-Leave-Experience-Report-Moms-at-Work-1.pdf)",
  Database = "Data from 2021 survey of 1090 mothers who had taken maternity leave in the last 10 years."
)

# Transpose the data frame
maternal <- t(maternal)

# Replace underscores with spaces in column names
colnames(maternal) <- gsub("_", " ", colnames(maternal))
# Create a table using kableExtra
  kable(maternal, format = "latex", caption = "Summary of findings -Parenthood Indicator") %>%

  column_spec(2, width = "29em")
```

This last indicator examines the challenges faced by new mothers returning to the workplace. Canada has legal provisions to ensure that maternal leave comes with benefits, and national statistics indicate that a substantial majority, 85.4% of new mothers with insurable employment, received maternity or parental benefits in 2018 according to the Employment Insurance Coverage Survey. However, it is crucial to consider the practical implementation of these laws, as their existence doesn't guarantee enforcement. A survey conducted among 1090 mothers who took maternity leave between 2011 and 2021 reveals that one in three women encountered discrimination upon their return to the workplace. This discrimination manifested in various forms, including the following:\
• Poor treatment and derogatory comments\
• Denied opportunities to develop professionally (projects, training, responsibility)\
• Laid off, restructured or demoted\
• Denied promotion, new role or advancement of earnings\
• Inadequate or denial of accommodations.

Furthermore, a striking 95% of the surveyed women did not receive any formal support during their maternity leave transition. Notably, 40% seriously contemplated resigning during the return-to-work process.

# Discussion

## Concluding remarks

In conclusion, the Canadian lens applied to the WBL Database provide a

## Rigged data and the World Bank

## Future research

@@

::: {#refs}

@@

# Appendix {.unnumbered}
